Linting Linearity in Core/System FC

### 2023-01-19 report.tex
texing help

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}

\and

  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}

    \and
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}

    \and

    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}

    \and

    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}

    \and

    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

    \and

    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}

    \and

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\star I$)]
    { }
    {\Delta \vdash \star : \star}

    \and

    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}

\end{mathpar}

\begin{mathpar}
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}

    \and

    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\end{mathpar}

@@

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input. A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \textbf{head} function by expressing
that the size of the list must be at least one, at the type level.

% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
% include recursive ADTs
\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
which users can define data types as they would common recursive algebraic data
types, but that allows explicit type signatures for the data constructors where
the result type may differ in the type parameters across branches. Pattern
matching against GADTs can introduce local type refinements, that is, pattern
matching against certain data constructor will refine the type information known
for typechecking in individual case alternatives. We develop the length-indexed
lists example without discussing the type system and type inference details of
GADTs which we explore later in~\ref{related-work-gadts}.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
possible different data type constructors being parametrized over different type
parameters as we do with \textbf{Vec}'s size parameter being different depending
on the constructor that constructs the list.
    To define the safe \textbf{head} function, we must specify the type of the
input list taking into account that the size must not be zero. To that effect,
the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.

>>>>

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}
\and
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

\qquad
    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
  \]

@@

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input.
\todo[inline]{When applied to the empty list, ...}

A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textbf{NonEmpty} is a list that can not be empty).

\todo[inline]{Leveraging Haskell's more advanced features, we can
  .... }

A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \textbf{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
which users can define data types as they would common recursive algebraic data
types, but that allows explicit type signatures for the data constructors where
the result type may differ in the type parameters across branches. Pattern
matching against GADTs can introduce local type refinements, that is, pattern
matching against certain data constructor will refine the type information known
for typechecking in individual case alternatives. We develop the length-indexed
lists example without discussing the type system and type inference details of
GADTs which we explore later in~\ref{related-work-gadts}.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \texttt{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
possible different data type constructors being parametrized over different type
parameters as we do with \textbf{Vec}'s size parameter being different depending
on the constructor that constructs the list.
    To define the safe \textbf{head} function, we must specify the type of the
input list taking into account that the size must not be zero. To that effect,
the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.
